Problem of Step 4 (debias) of fGFPCA on NHANES data

### 2024-01-19 Docs/NhanesApplExample.Rmd
summary stats for PC score estimates

library(here)
library(tidyverse)
library(ggpubr)
library(refund)
library(lme4)
library(ROCR)
library(GLMMadaptive)
library(kableExtra)
library(mvtnorm)
library(mgcv)
library(splines)
library(LaplacesDemon)
theme_set(theme_minimal())
```

@@

Follow up on the email yesterday, we would like to see what would happen if we only take the first **three** PC functions.

@@

- Eigenvalues: `r fpca_mod$evalues`
- Proportion of variance explained with four PCs: `r fpca_mod$pve`

@@

Again, here we use the three PC functions in the random effect part.

```{r revaluation}
df_phi <- data.frame(sind = 1:J, df_phi)
colnames(df_phi) <- c("sind", paste0("phi", 1:K))
train_df <- train_df %>% 
  left_join(df_phi, by = "sind")
train_df$id <- as.factor(train_df$id)

t1 <- Sys.time()
debias_glmm <- bam(Y ~ s(sind, bs="cc")+
                     s(id, by=phi1, bs="re")+
                     s(id, by=phi2, bs="re")+
                     s(id, by=phi3, bs="re")+
                     s(id, by=phi4, bs="re"), 
                   family = binomial, 
                   data=train_df, 
                   method = "fREML",
                   discrete = TRUE)
t2 <- Sys.time()
# t2-t1 # 7.5 minutes
```

It seems that the eigenvalues are still losing their decreasing nature. The eigenvalue corresponding to PC2 is still smaller than PC3. Specifically

- The smoothing parameters from the mgcv::bam model: `r debias_glmm$sp[2:5]`
- The re-evaluated eigenvalues, which are inverse of the smoothing parameters above: `r 1/debias_glmm$sp[2:5]`
- The **biased** eigenvalues from FPCA (step 3): `r fpca_mod$evalues[1:4]`

Also, the proportion of these two set of estimates is not constant, as below:

```{r, class.source = 'fold-show'}
# proportion of biased eigenvalues to re-evaluated eigenvalues
fpca_mod$evalues/(1/debias_glmm$sp[2:5])
```

### Percentage of variance explained by each PC

We now that the variance of individual-specific random effects, in theory, is:

@@

Here I'd like to compare the variantion for each PC calculated using estimated from the **biased** FPCA estimates and the supposedly **debiased** from the dibased GLMM model.

@@

bind_rows(
  df_pc_fpca %>% select(t, starts_with("var"), est),
  df_pc_glmm %>% select(sind, starts_with("var"), est) %>% rename(t=sind)
) %>% 
  rename(PC1=var1, PC2=var2, PC3=var3, PC4=var4) %>%
  pivot_longer(2:5, names_to = "PC", values_to = "variance") %>%
  ggplot()+
  geom_line(aes(x=t, y=variance, col=est, group=est))+
  facet_wrap(~PC)+
  labs(title = "Time-varying variation of each PC")

@@

I am also curious about how the individual scores change after debias

@@

@@

```{r}
bind_rows(df_score_fpca, df_score_glmm) %>%
  ggplot(aes(x=est, y=score, col=est))+
  geom_boxplot()+
  geom_jitter(size = 0.2, alpha=0.5)+
  facet_wrap(~PC, scales = "free_y")+
  labs(title = "Score of each PC")

@@

>>>>

library(here)
library(tidyverse)
library(ggpubr)
library(refund)
library(lme4)
library(ROCR)
library(GLMMadaptive)
library(kableExtra)
library(mvtnorm)
library(mgcv)
library(splines)
library(LaplacesDemon)
library(arsenal)
theme_set(theme_minimal())
```

@@

@@

- Proportion of variance explained with four PCs: `r round(fpca_mod$pve, 3)`

@@

```{r revaluation}
df_phi <- data.frame(sind = 1:J, df_phi)
colnames(df_phi) <- c("sind", paste0("phi", 1:K))
train_df <- train_df %>% 
  left_join(df_phi, by = "sind")
train_df$id <- as.factor(train_df$id)
```

```{r, class.source='fold-show'}
t1 <- Sys.time()
debias_glmm <- bam(Y ~ s(sind, bs="cr")+
                     s(id, by=phi1, bs="re")+
                     s(id, by=phi2, bs="re")+
                     s(id, by=phi3, bs="re")+
                     s(id, by=phi4, bs="re"), 
                   family = binomial, 
                   data=train_df, 
                   method = "fREML",
                   discrete = TRUE)
t2 <- Sys.time()
# t2-t1 # less than 7.5 minutes
```

It seems that the re-evaluated eigenvalues lost their decreasing nature. The eigenvalue corresponding to PC2 is smaller than PC3. Specifically

- The smoothing parameters from the mgcv::bam model: `r round(debias_glmm$sp[2:5], 4)`

- The re-evaluated eigenvalues, which are inverse of the smoothing parameters above: `r round(1/debias_glmm$sp[2:5], 3)`
- The **biased** eigenvalues from FPCA (step 3): `r round(fpca_mod$evalues[1:4], 3)`

Below I try to investigate the cause of this problem.

### Variance explained by each PC

The variance of individual-specific random effects, in theory, is:

@@

Here I'd like to compare the variation from each PC calculated using estimated from the **biased** FPCA estimates (binned grid) and the supposedly **debiased** from the dibased GLMM model (original grid).

@@

bind_rows(
  df_pc_fpca %>% select(t, starts_with("var"), est),
  df_pc_glmm %>% select(sind, starts_with("var"), est) %>% rename(t=sind)
) %>% 
  rename(PC1=var1, PC2=var2, PC3=var3, PC4=var4) %>%
  pivot_longer(2:5, names_to = "PC", values_to = "variance") %>%
  ggplot()+
  geom_line(aes(x=t, y=variance, col=est, group=est))+
  facet_wrap(~PC)+
  labs(title = "Time-varying variation of each PC", x="Time", y = "Variation",
       col = "")

@@

### Estimates of score

I am also curious about how the individual score estimates change after debias

@@

df_score <- bind_rows(df_score_fpca, df_score_glmm) %>%
  mutate(PC = gsub("X", "PC", PC)) 
```

```{r score_box}
df_score %>%
  ggplot(aes(x=est, y=score, col=est))+
  geom_boxplot()+
  geom_jitter(size = 0.2, alpha=0.5)+
  facet_wrap(~PC, scales = "free_y")+
  labs(title = "Score estimates of each PC", x = "", y="Score estimates",
       col = "")

@@

```{r score_hist}
df_score %>% 
  ggplot()+
  geom_histogram(aes(x=score), bins = 100)+
  facet_grid(rows = vars(PC), cols = vars(est), scales = "free")+
  labs(title = "Distribution of score estimates", x = "Score estimates",
       y= "")

@@

From figures above, the score estimates on the binned grid (FPCA) of the 3rd PC is quite skewed negatively (and the 2nd PC slightly positively?)

Below are summary statistics

```{r score_sum_stat}
df_score %>%
  pivot_wider(names_from = PC, values_from = score) %>%
  tableby(est ~ PC1 + PC2 + PC3 + PC4, data = .,
          control = tableby.control(total = F, 
                                    test = F)) %>% 
  summary() %>% 
  kable(table.attr = "style = \"color: black;\"", escape =F) %>% 
  kable_styling(full_width = F)
